---
title: "Assignment-3-Data-Manipulation-in-R"
author: "Nwaneri Tochi , Mr. Charles Nworu"
date: "2025-04-29"
output: html_document
---


```{r,load_library, include=FALSE}
library(dplyr)
library(readxl)

# Load Excel document
my_data <- read_excel("Countries.xlsx")


```


# 1. How does the GDP per Capita in 2000 compare between countries in Asia and countries in Africa?
```{r}
data_2000 <- my_data%>%
  filter(Year == 2000)

asia_2000 <- my_data%>%
  filter (`Continent Name`== "asia")%>%
select(`Country Code`,`Country Name`,`GDP Per Capita`)

africa_2000 <- my_data%>%
  filter(`Continent Name` == "asia")%>%
  select(`Country Code`,`Country Name`, `GDP Per Capita`)

joined <- inner_join(
  asia_2000,africa_2000, by = 'Country Code',
  suffix = c("Asia","Africa")
)

comparison <- joined%>%
  mutate(GDP_Difference = `GDP Per CapitaAsia`-`GDP Per CapitaAfrica`)%>%
  arrange(desc(abs(GDP_Difference)))

comparison
```




# 2. Can we enrich a dataset of country codes from 2020 with full details from this dataset using right_join()?
```{r}
#Get all unique country codes from 2020

codes_2020 <- my_data%>%
filter(Year == 2020)%>%
  select(`Country Code`)%>%
  distinct()%>%
  mutate(Year= 2020)

#Right join to enrich
enriched_data <- right_join(
  my_data,codes_2020, by = c("Year","Country Code")
)
head(enriched_data)
```

# 3.   Which countries in 2000 reported data for Health Expenditure but not for Export values?
```{r}
#Get Countries with Health Expenditure data

health_2000 <- data_2000%>%
  filter(!is.na(`Health Expenditure`))%>%
  select(`Country Name`,`Country Code`)

#Get countries with Export data
export_2000 <- data_2000%>%
  filter(!is.na(Export))%>%
  select(`Country Name`, `Country Code`)

#Countries with Health Expenditure
has_health <- semi_join(data_2000,health_2000,by = "Country Code")

#Exclude countries that also reported Export values 
final_result <- anti_join(has_health,export_2000,by = "Country Code")

#View result
final_result%>%select("Country Name","Country Code","Health Expenditure")
```


# 4.Which countries appear in both the 2000 and 2010 datasets?
```{r}
# Extract county Names for 2000 and 2010
countries_2000 <- my_data%>%
  filter(Year == 2000)%>%
  pull(`Country Name`)%>%
  unique()

countries_2010 <- my_data%>%
  filter(Year == 2010)%>%
  pull(`Country Name`)%>%
  unique()

#Find the intersection of both lists
common_countries <- intersect(countries_2000,countries_2010)
common_countries
```

# 5.Within each continent, which country had the highest GDP per Capita  in 2010?
```{r}
top_gdp_per_capita_rate <- my_data%>%
  filter(Year==2010)%>%
  group_by(`Continent Name`)%>%
  filter(rank(desc(`GDP Per Capita`))==1)%>%
ungroup()
top_gdp_per_capita_rate
```


# 6. For a selected country (e.g., Nigeria), how did the GDP Per Capita change year-over-year?
```{r}
nigeria_gdp_change <- my_data%>%
  filter(`Country Name` == "Nigeria")%>%
  arrange(Year)%>%
  mutate(
    gdp_lag = lag(`GDP Per Capita`),
    gdp_change = `GDP Per Capita` -`gdp_lag`
  )


nigeria_gdp_change
```

# 7.  What is the cumulative Health Expenditure of each country from 2000 to 2005?
```{r}
cumulative_health <- my_data %>%
  filter (Year >= 2000, Year <=2005)%>%
  arrange(`Country Name`,`Year`)%>%
  group_by(`Country Name`)%>%
  mutate(cum_health = cumsum(`Health Expenditure`))%>%
  ungroup()

cumulative_health
```



# 8. What is the mean of all numeric economic indicators (like GDP,Education Expenditure) for each country in 2015?  
```{r}
economic_mean_2015 <- my_data%>%
filter(Year == 2015)%>%
group_by(`Country Name`)%>%
summarise(across(where(is.numeric),mean,na.rm = T))

economic_mean_2015
```


# 9. Which rows have at least one economic indicator (e.g., Export, Import,Health, Education) greater than 15% of GDP?
```{r}
high_econ_indicators <- my_data%>%
  filter(if_any(ends_with("% GDP"), ~. > 15))

high_econ_indicators
```


