---
title: "Assignment 2 _ Data Manipulation In R"
author: "Nwaneri Tochi , Mr. Charles Nworu"
date: "2025-04-22"
output: html_document
---


```{r setup, include=FALSE}
# load libraries
library(dplyr)
library(readxl)
```


```{r} 
# load the dataset
data <- read_excel("C:/Users/Owner/Documents/countries.xlsx")


#View first few rows
head(data)
```

1a. Countries in 2010 with GDP > $500 billion
```{r}
data_2010_gdp <- data %>%
  filter(Year == 2010, GDP > 500e9)
data_2010_gdp
```

1b. African countries with population density > 100
```{r}
africa_density <- data %>%
  filter(`Continent Name` =="Africa", `Population Density` > 100 )
africa_density
```

1c. Agriculture  < 10% and Health expenditure > 8%
```{r}
agri_health <- data %>%
  filter(`Agriculture (% GDP)` < 10, `Health Expenditure (% GDP)` > 8)
agri_health
```

1d. List the countries from Asia where Education Expenditure was not missing in the year 2005
```{r}
asia_edu_2005 <- data %>%
  filter(`Continent Name` == "Asia", Year == "2005", !is.na(`Education Expenditure`))
asia_edu_2005
```

2a.	Select only the Country Name, Year, and GDP columns.
```{r}
selected_cols <-  data %>%
  select(`Country Name`, Year, GDP)
selected_cols
```
2b.	Use a helper function to select all columns that start with "Health".
```{r}
health_cols <- data%>%
  select(starts_with("Health"))
health_cols
```
2c.	Select a range of columns from GDP to Import (% GDP) to analyze economic indicators
```{r}
economic_indicators <- data%>%
  select(GDP:`Import (% GDP)`)
economic_indicators
```
3a.	Create a new column GDP in Billions by dividing GDP by 1e9.
```{r}
data <- data %>%
  mutate('GDP Billions' = GDP / 1e9)
head(data)

```
3b.	Add a column called High Population to label entries where the population is above 100 million.
```{r}
data <- data %>%
  mutate('High Population' = Population > 100e6)
head(data)
```
3c.	Use case_when() to categorize countries into "Low", "Medium", and "High" GDP per capita based on thresholds of 1000 and 10000.
```{r}
data <- data%>%
  mutate('GDP CATEGORY' = case_when(
    `GDP Per Capita` < 1000 ~ "Low",
    `GDP Per Capita` >= 1000 & `GDP Per Capita` < 10000 ~ "Medium",
    `GDP Per Capita` >= 10000 ~ "High"
  ) )

head(data)

```
I made a mistake by adding two columns of GDP Billions; GDP_Billions & GDP Billions. I want to get rid of the one with the underscore
```{r}
data <- data %>%
  mutate('GDP_Billions' = NULL)
head(data)
```
4a.	Sort the dataset by GDP in descending order to find the highest GDP countries
```{r}
sorted_gdp <- data%>%
  arrange(desc(GDP))
head(sorted_gdp)
```
4b.	Arrange countries within the same continent in ascending order of Population Density.
```{r}
sorted_by_continent_pop <- data%>%
  arrange(`Continent Name`,`Population Density`)
head(sorted_by_continent_pop)
```
5a.	What is the average GDP per capita for each continent?
```{r}
avg_gdp_per_continent <- data %>%
  group_by(`Continent Name`)%>%
  summarise(Average_GDP_Per_Capita = mean(`GDP Per Capita`, na.rm = T))
avg_gdp_per_continent
```
5b.	How many records are there for each country in the dataset?
```{r}
data%>%
  group_by(`Country Name`) %>%
  summarise(record_count = n())
```
d.	Calculate the mean, minimum, and maximum of Health Expenditure (% GDP) grouped by Continent Name.\
```{r}
data %>%
  group_by(`Continent Name`)%>%
  summarise(
    Mean_health_Expenditure = mean(`Health Expenditure (% GDP)`, na.rm = T),
    Min_Health_Expenditure = min(`Health Expenditure (% GDP)`, na.rm = T ),
    Max_Health_Expenditure = max(`Health Expenditure (% GDP)`, na.rm =  T)
  )
head(data)
```
6a.	Chain together operations to filter only African countries from 2015, then select the Country Name, GDP, and Population, and arrange them in descending order of GDP.
```{r}
data%>%
  filter(`Continent Name` == "Africa", Year == 2015)%>%
  select(`Country Name`, GDP, Population)%>%
  arrange(desc(GDP))
head(data)
```
6b.	For each continent, calculate the average GDP Per Capita, but only for countries with Education Expenditure (% GDP) greater than 4%.
```{r}
data %>%
  filter(`Education Expenditure (% GDP)` > 4) %>%
  group_by(`Continent Name`) %>%
  summarise(Average_GDP_Per_Capita = mean(`GDP Per Capita`, na.rm = T))
```
6c.	Filter countries with a population over 50 million, create a column for GDP in billions, and show only the top 5 highest GDPs.
```{r}
data %>%
  filter(Population > 50e6) %>%
  mutate(`GDP Billions` = GDP / 1e9)%>%
  select(`Country Name`, Population, `GDP Billions`) %>%
  slice_head(n = 5)
head(data)
```
7a.	Rename GDP Per Capita to GDP_per_capita_USD and Country Name to country_name.
```{r}
data <- data %>%
  rename(
    `GDP_Per_Capita_USD` = `GDP Per Capita`,`Country_Name` = `Country Name`
  )
names(data)
```
7b.	Change Health Expenditure (% GDP) to health_pct_gdp for clarity in plotting
.
```{r}
data <- data %>%
  rename (
    health_pct_gdp = `Health Expenditure (% GDP)`
  )
head(data)
```
8a.	Move Year to the front of the dataset.
```{r}
data <- data %>%
  relocate (Year, .before = 1)
head(data)
  
```
8b.	Relocate Population and GDP to appear right after Country Name for better visibility during analysis.
```{r}
data <- data%>%
  relocate(Population,GDP, .after = Country_Name)
head(data)
```
8c.	Reorder Continent Name and Country Code to appear at the end
```{r}
data <- data%>%
  relocate(`Continent Name`, `Country Code`, .after = last_col())
head(data)
```
9a.	How many unique countries are in the dataset?
```{r}
n_distinct(data$`Continent Name`)
```
9b.	Extract distinct combinations of Country Name and Continent Name.
```{r}
data %>%
  distinct(Country_Name, `Continent Name`)
```


9c.	How many distinct years of data are available?
```{r}
n_distinct(data$Year)
```
10a.	Display the first 10 rows of countries sorted by Population Density for 2020.
```{r}
data %>%
  filter(Year == 2020) %>%
  arrange(desc(`Population Density`)) %>%
  slice_head(n = 10)
```
10b. Show the last 5 entries of GDP figures for Nigeria.
```{r}
data %>%
  filter(Country_Name == "Nigeria") %>%
  arrange(Year) %>%
  slice_tail(n = 5)

```
10c.	Slice to extract the 3rd to 7th rows from data on South American countries.
```{r}
data %>% 
  filter(`Continent Name`== "South America") %>%
  slice(3:7)
```
11a comparing the performance between base R and Dplyr for 'Which countries in 2010 had a GDP greater than $500 billion?'
```{r}
time_base_a <- system.time({
  base_a <- data[data$Year == 2010 & data$GDP > 500e9, c("Country_Name", "GDP")]
})
time_base_a



time_dplyr_a <- system.time({
  dplyr_a <- data %>%
    filter(Year == 2010, GDP > 500e9) %>%
    select(Country_Name, GDP)
})
time_dplyr_a
```
11b. comparing the performance between base R and Dplyr for 'Filter the data to show only African countries with a population density greater than 100'.
```{r}
time_base_b <- system.time({
  base_b <- data[data$`Continent Name`== "Africa" & data$`Population Density` > 100, c("Country_Name", "Population Density")]
})
 time_base_b
 
 
 time_dplyr_b <- system.time({
  dplyr_b <- data %>%
    filter(`Continent Name`== "Africa", `Population Density` > 100) %>%
    select(Country_Name, `Population Density`)
})
time_dplyr_b
```
11c. comparing the performance between base R and Dplyr for 'What countries had both an agriculture contribution of less than 10% and health expenditure above 8% of GDP?'
```{r}
time_base_c <- system.time({
  base_c <- data[data$`Agriculture (% GDP)` < 10 & data$`health_pct_gdp` > 8,
                 c("Country_Name", "Agriculture (% GDP)", "health_pct_gdp")]
})
time_base_c

time_dplyr_c <- system.time({
  dplyr_c <- data %>%
    filter(`Agriculture (% GDP)` < 10, `health_pct_gdp` > 8) %>%
    select(Country_Name, `Agriculture (% GDP)`, `health_pct_gdp`)
})
time_dplyr_c

```
11d.comparing the performance between base R and Dplyr for 'List the countries from Asia where Education Expenditure was not missing in the year 2005.'
```{r}
time_base_d <- system.time({
  base_d <- data[data$`Continent Name` == "Asia" & data$Year == 2005 & !is.na(data$`Education Expenditure (% GDP)`),
                 c("Country_Name", "Education Expenditure (% GDP)")]
})
time_base_d

time_dplyr_d <- system.time({
  dplyr_d <- data %>%
    filter(`Continent Name` == "Asia", Year == 2005, !is.na(`Education Expenditure (% GDP)`)) %>%
    select(Country_Name, `Education Expenditure (% GDP)`)
})
time_dplyr_d


```
```{r}
head(data)

```





